Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on

### 2011-06-22 paper.tex
Halfway through final edits

\author{
  Noel M O'Boyle\correspondingauthor$^1$%
  \email{Noel M O'Boyle\correspondingauthor - baoilleach@gmail.com}
  \and
  Samuel E Adams$^2$%
  \email{Samuel E Adams - sea36@cam.ac.uk}
  \and
  Jonathan Alvarsson$^3$%
  \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
  \and
  Richard L Apodaca$^{4}$%
  \email{Richard L Apodaca - rapodaca@metamolecular.com}
  \and
  Jean-Claude Bradley$^5$%
  \email{Jean-Claude Bradley - bradlejc@drexel.edu}
  \and
  Igor V Filippov$^6$%
  \email{Igor V Filippov - igorf@helix.nih.gov}
  \and
  Rajarshi Guha$^7$%
  \email{Rajarshi Guha - guhar@mail.nih.gov}
  \and
  Robert M Hanson$^8$%
  \email{Robert M Hanson - hansonr@stolaf.edu}
  \and
  Marcus D Hanwell$^{9}$%
  \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
  \and
  Geoffrey R Hutchison$^{10}$%
  \email{Geoffrey R Hutchison - geoffh@pitt.edu}
  \and
  Andrew SID Lang$^{11}$%
  \email{Andrew SID Lang - alang@oru.edu}
  \and
  Karol M Langner$^{12}$%
  \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
  \and
  Daniel M Lowe$^2$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
  \and
  Dmitry Pavlov$^{13}$%
  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
  \and
  Ola Spjuth$^3$%
  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
  \and
  Christoph Steinbeck$^{14}$%
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
  \and
  Adam L Tenderholt$^{15}$%
  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
  \and
  Kevin J Theisen$^{16}$%
  \email{Kevin J Theisen - kevin@ichemlabs.com}
  \and
  Egon L Willighagen$^{17}$%
  \email{Egon L Willighagen - egon.willighagen@ki.se}
  \and
  Peter Murray-Rust$^2$%
  \email{Peter Murray-Rust - pm286@cam.ac.uk}
}

@@

\address{%
    \iid(1)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
    \iid(2)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
    \iid(4)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
    \iid(5)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
    \iid(6)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
Inc., NCI-Frederick, Frederick, MD 21702, USA \\
    \iid(7)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
    \iid(8)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
    \iid(9)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
    \iid(11)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
    \iid(12)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
    \iid(13)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
    \iid(14)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
    \iid(15)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
    \iid(16)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
    \iid(17)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden 
}%

@@

A final example of second-generation Blue Obelisk software is the 
AMBIT2~\cite{Jeliazkova2011} software, a GUI that facilitates registration
of chemicals for the REACH EU directive on toxicity, and which is
based on the CDK.

@@

Despite its widespread use, a formal
definition of the language did not exist beyond Daylight's SMILES
Theory Manual and tutorials. This caused some confusion in the
implementation and interpretation of corner cases, for example the
handling of cis/trans bond symbols at ring closures. In 2007, Craig
James (eMolecules) initiated work on the OpenSMILES specification, a
complete specification of the SMILES language as an Open Standard
developed through a community process. The specification is largely
complete and contains guidelines on reading SMILES, a formal
grammar, recommendations on standard forms when writing SMILES, as
well as proposed extensions.

@@

The Blue Obelisk has created a repository of key chemical
data in a machine-readable format~\cite{BODR}.
The BODR focuses on data that is commonly required for
chemistry software, and where there is a need to ensure that values
are standard between codes. Examples are atomic
masses and conversions between physical constants. These data
can be used by others for any purpose (for example, for entry into
Wikipedia or use in in-house software), and should lead
to an enhancement in the quality of community reference data.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Blue Obelisk Open Source Software projects discussed in the text}
    \par \mbox{}
    \par
    \mbox{
     \begin{tabular}{|c|c|c|}
       \hline \textbf{Name} & \textbf{Website}  \\ \hline
        \multicolumn{2}{|c|}{\textbf{CML Tools}} \\ \hline
        \textbf{CMLXOM} & https://bitbucket.org/wwmm/cmlxom/ \\ \hline
        \textbf{JUMBO} & http://sourceforge.net/projects/cml/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
        \textbf{Chemistry Development Kit (CDK)} & http://cdk.sf.net \\ \hline
        \textbf{Cinfony} & http://cinfony.googlecode.com \\ \hline
        \textbf{Indigo} & http://ggasoftware.com/opensource/indigo \\ \hline
        \textbf{JOELib} & http://sf.net/projects/joelib \\ \hline
        \textbf{Open Babel} & http://openbabel.org \\ \hline
        \textbf{RDKit} & http://rdkit.org \\ \hline
        \multicolumn{2}{|c|}{\textbf{Web Applications}} \\ \hline
        \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com \\ \hline
        \textbf{gChem} & http://metamolecular.com/gchem/ \\ \hline
        \textbf{Jmol} & http://jmol.org \\ \hline
        \multicolumn{2}{|c|}{\textbf{Integration}} \\ \hline
        \textbf{Bioclipse} & http://www.bioclipse.net \\ \hline
        \textbf{CDK-Taverna} & http://cdktaverna.wordpress.com \\ \hline
        \textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Interconversion}} \\ \hline
        \textbf{CIFXOM} \cite{DayEtAl2011} & https://bitbucket.org/wwmm/cifxom/  \\ \hline
        \textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ \\ \hline
        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk \\ \hline
        \textbf{OSRA} & http://osra.sf.net \\ \hline
        \multicolumn{2}{|c|}{\textbf{Structure Databases}} \\ \hline
        \textbf{Bingo}  & http://ggasoftware.com/opensource/bingo \\ \hline
        \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound \\ \hline
        \textbf{Mychem}  & http://mychem.sf.net \\ \hline
        \textbf{OrChem} & http://orchem.sf.net \\ \hline
        \textbf{pgchem}  & http://pgfoundry.org/projects/pgchem/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Text mining}} \\ \hline
        \textbf{ChemicalTagger} \cite{HawizyEtAl2011} & http://chemicaltagger.ch.cam.ac.uk/ \\ \hline
        \textbf{OSCAR4} & https://bitbucket.org/wwmm/oscar4/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Computational Chemistry}} \\ \hline
        \textbf{Avogadro} & http://avogadro.openmolecules.net \\ \hline
        \textbf{cclib}  & http://cclib.sf.net \\ \hline
        \textbf{GaussSum}  & http://gausssum.sf.net \\ \hline
        \textbf{QMForge}  & http://qmforge.sf.net \\ \hline
        \multicolumn{2}{|c|}{\textbf{Computational Drug Design}} \\ \hline
        \textbf{Confab}~\cite{Confab} & http://confab.googlecode.com \\ \hline
        \textbf{Pharao} & http://silicos.be/download \\ \hline
        \textbf{Piramid} & http://silicos.be/download \\ \hline
        \textbf{Sieve} & http://silicos.be/download \\ \hline
        \textbf{Stripper} & http://silicos.be/download \\ \hline
        \multicolumn{2}{|c|}{\textbf{Other Applications}} \\ \hline
        \textbf{AMBIT2}  & http://ambit.sf.net \\ \hline
        \textbf{Brunn}  & http://brunn.sf.net \\ \hline
        \textbf{XtalOpt}  & http://xtalopt.openmolecules.net \\ \hline
        \hline
       \end{tabular}
      }

>>>>

\author{
  Noel M O'Boyle\correspondingauthor$^1$%
  \email{Noel M O'Boyle\correspondingauthor - baoilleach@gmail.com}
  \and
  Rajarshi Guha$^2$%
  \email{Rajarshi Guha - guhar@mail.nih.gov}
  \and
  Egon L Willighagen$^{3}$%
  \email{Egon L Willighagen - egon.willighagen@ki.se}
  \and
  Samuel E Adams$^4$%
  \email{Samuel E Adams - sea36@cam.ac.uk}
  \and
  Jonathan Alvarsson$^5$%
  \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
  \and
  Richard L Apodaca$^{6}$%
  \email{Richard L Apodaca - rapodaca@metamolecular.com}
  \and
  Jean-Claude Bradley$^7$%
  \email{Jean-Claude Bradley - bradlejc@drexel.edu}
  \and
  Igor V Filippov$^8$%
  \email{Igor V Filippov - igorf@helix.nih.gov}
  \and
  Robert M Hanson$^9$%
  \email{Robert M Hanson - hansonr@stolaf.edu}
  \and
  Marcus D Hanwell$^{10}$%
  \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
  \and
  Geoffrey R Hutchison$^{11}$%
  \email{Geoffrey R Hutchison - geoffh@pitt.edu}
  \and
  Craig A James$^{12}$%
  \email{Craig A James - cjames@emolecules.com}
  \and
  Nina Jeliazkova$^{13}$%
  \email{Nina Jeliazkova - jeliazkova.nina@gmail.com}
  \and
  Andrew SID Lang$^{14}$%
  \email{Andrew SID Lang - alang@oru.edu}
  \and
  Karol M Langner$^{15}$%
  \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
  \and
  David C Lonie$^{16}$%
  \email{David C Lonie - loniedavid@gmail.com}
  \and
  Daniel M Lowe$^4$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
  \and
  J\'{e}r\^{o}me Pansanel$^{17}$%
  \email{J\'{e}r\^{o}me Pansanel - j.pansanel@pansanel.net}
  \and
  Dmitry Pavlov$^{18}$%
  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
  \and
  Ola Spjuth$^{5}$%
  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
  \and
  Christoph Steinbeck$^{19}$%
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
  \and
  Adam L Tenderholt$^{20}$%
  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
  \and
  Kevin J Theisen$^{21}$%
  \email{Kevin J Theisen - kevin@ichemlabs.com}
  \and
  Peter Murray-Rust$^4$%
  \email{Peter Murray-Rust - pm286@cam.ac.uk}
}

@@

\address{%
    \iid(1)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
    \iid(2)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden \\
    \iid(4)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
    \iid(5)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
    \iid(6)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
    \iid(7)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
    \iid(8)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
Inc., NCI-Frederick, Frederick, MD 21702, USA \\
    \iid(9)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
    \iid(10)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
    \iid(11)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
    \iid(12)eMolecules Inc., 380 Stevens Ave., Solana Beach, California 92075, USA \\
    \iid(13)Ideaconsult Ltd., 4.A.Kanchev str., Sofia 1000, Bulgaria \\
    \iid(14)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
    \iid(15)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
    \iid(16)Department of Chemistry, State University of New York at Buffalo, Buffalo, NY 14260-3000, USA \\
    \iid(17)Universit\'{e} de Strasbourg, IPHC, CNRS, UMR7178, 23 rue du Loess 67037, Strasbourg, France \\
    \iid(18)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
    \iid(19)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
    \iid(20)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
    \iid(21)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA 
}%

@@

A final example of second-generation Blue Obelisk software is the 
AMBIT2~\cite{Jeliazkova2011,InSilicoTox_2010} software, which was developed to facilitate registration of chemicals for the REACH EU directive, and is based on the CDK. It was distributed initially as a standalone Java Swing GUI, and more recently as downloadable web application archive, offering a web services interface to a searchable chemical structures database. Also integrated are descriptor calculations, as well as the ability to run and build predictive models, including modules of the open source Toxtree~\cite{WebToxTree,InSilicoTox_2010,Toxtree} software for toxicity prediction.

@@

Despite its widespread use, a formal
definition of the language did not exist beyond Daylight's SMILES
Theory Manual and tutorials. This caused some confusion in the
implementation and interpretation of corner cases, for example the
handling of cis/trans bond symbols at ring closures. In 2007, Craig
James (eMolecules) initiated work on the OpenSMILES specification~\cite{WebOpenSMILES}, a
complete specification of the SMILES language as an Open Standard
developed through a community process. The specification is largely
complete and contains guidelines on reading SMILES, a formal
grammar, recommendations on standard forms when writing SMILES, as
well as proposed extensions.

@@

The Blue Obelisk has created a repository of key chemical
data in a machine-readable format~\cite{BODR}.
The BODR focuses on data that is commonly required for
chemistry software, and where there is a need to ensure that values
are standard between codes. Examples are atomic
masses and conversions between physical constants. These data
can be used by others for any purpose (for example, for entry into
Wikipedia or use in in-house software), and should lead
to an enhancement in the quality of community reference data.
The Blue Obelisk provides also a complementary project, the Chemical
Structure Repository~\cite{BODR}. It aims to provide 3D coordinates, InChi code and several
physico-chemical descriptors for a set of 570 organic compounds.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Blue Obelisk Open Source Software projects discussed in the text}
    \par \mbox{}
    \par
    \mbox{
     \begin{tabular}{|c|c|c|}
       \hline \textbf{Name} & \textbf{Website}  \\ \hline
        \multicolumn{2}{|c|}{\textbf{CML Tools}} \\ \hline
        \textbf{CMLXOM} & https://bitbucket.org/wwmm/cmlxom/ \\ \hline
        \textbf{JUMBO} & http://sourceforge.net/projects/cml/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
        \textbf{Chemistry Development Kit (CDK)} & http://cdk.sf.net \\ \hline
        \textbf{Cinfony} & http://cinfony.googlecode.com \\ \hline
        \textbf{Indigo} & http://ggasoftware.com/opensource/indigo \\ \hline
        \textbf{JOELib} & http://sf.net/projects/joelib \\ \hline
        \textbf{Open Babel} & http://openbabel.org \\ \hline
        \textbf{RDKit} & http://rdkit.org \\ \hline
        \multicolumn{2}{|c|}{\textbf{Web Applications}} \\ \hline
        \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com \\ \hline
        \textbf{gChem} & http://metamolecular.com/gchem/ \\ \hline
        \textbf{Jmol} & http://jmol.org \\ \hline
        \multicolumn{2}{|c|}{\textbf{Integration}} \\ \hline
        \textbf{Bioclipse} & http://www.bioclipse.net \\ \hline
        \textbf{CDK-Taverna} & http://cdktaverna.wordpress.com \\ \hline
        \textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Interconversion}} \\ \hline
        \textbf{CIFXOM} \cite{DayEtAl2011} & https://bitbucket.org/wwmm/cifxom/  \\ \hline
        \textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ \\ \hline
        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk \\ \hline
        \textbf{OSRA} & http://osra.sf.net \\ \hline
        \multicolumn{2}{|c|}{\textbf{Structure Databases}} \\ \hline
        \textbf{Bingo}  & http://ggasoftware.com/opensource/bingo \\ \hline
        \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound \\ \hline
        \textbf{Mychem}  & http://mychem.sf.net \\ \hline
        \textbf{OrChem} & http://orchem.sf.net \\ \hline
        \textbf{pgchem}  & http://pgfoundry.org/projects/pgchem/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Text mining}} \\ \hline
        \textbf{ChemicalTagger} \cite{HawizyEtAl2011} & http://chemicaltagger.ch.cam.ac.uk/ \\ \hline
        \textbf{OSCAR4} & https://bitbucket.org/wwmm/oscar4/ \\ \hline
        \multicolumn{2}{|c|}{\textbf{Computational Chemistry}} \\ \hline
        \textbf{Avogadro} & http://avogadro.openmolecules.net \\ \hline
        \textbf{cclib}  & http://cclib.sf.net \\ \hline
        \textbf{GaussSum}  & http://gausssum.sf.net \\ \hline
        \textbf{QMForge}  & http://qmforge.sf.net \\ \hline
        \multicolumn{2}{|c|}{\textbf{Computational Drug Design}} \\ \hline
        \textbf{Confab}~\cite{Confab} & http://confab.googlecode.com \\ \hline
        \textbf{Pharao} & http://silicos.be/download \\ \hline
        \textbf{Piramid} & http://silicos.be/download \\ \hline
        \textbf{Sieve} & http://silicos.be/download \\ \hline
        \textbf{Stripper} & http://silicos.be/download \\ \hline
        \multicolumn{2}{|c|}{\textbf{Other Applications}} \\ \hline
        \textbf{AMBIT2}  & http://ambit.sf.net \\ \hline
        \textbf{Brunn}  & http://brunn.sf.net \\ \hline
        \textbf{Toxtree}  & http://toxtree.sf.net \\ \hline
        \textbf{XtalOpt}  & http://xtalopt.openmolecules.net \\ \hline
        \hline
       \end{tabular}
      }
